---
title: "week08"
author: "Leon Tran"
date: "2023-11-08"
output:
  word_document: default
  pdf_document: default
---

```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
wisc.df
```

```{r}
wisc.data <- wisc.df[,-1]
wisc.data
```
```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
```

Q1. How many observations are in this dataset?
```{r}
observations <- nrow(wisc.df)
observations
```

Q2. How many of the observations have a malignant diagnosis?
```{r}
numdiagnosis <- table(diagnosis)
malignant <- numdiagnosis["M"]
malignant
```

Q3. How many variables/features in the data are suffixed with _mean?
```{r}
meancolumns <- grep("_mean$", names(wisc.df))
nummeancolumns <- length(meancolumns)
nummeancolumns
```

```{r}
colMeans(wisc.data)
apply(wisc.data,2,sd)
wisc.pr <- prcomp(wisc.data, scale = TRUE)
summary(wisc.pr)
```

Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
```{r}
x <- wisc.pr$sdev[1]^2 / sum(wisc.pr$sdev^2)
x
```

Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
```{r}
cumulative <- cumsum((wisc.pr$sdev^2) / sum(wisc.pr$sdev^2))
percent70 <- which.max(cumulative >= 0.70)
percent70
```

Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
```{r}
percent90 <- which.max(cumulative >= 0.90)
percent90
```

```{r}
biplot(wisc.pr)
```
Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?
- The plot is very messy and hard to observe. The data is cluttered and overlaps so no patterns or trends can be determined. 
```{r}
plot(wisc.pr$x[, 1], wisc.pr$x[, 2], col = diagnosis, xlab = "PC1", ylab = "PC2")
```

Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?
-These plots are very similar in that the data points are cluttered and overlap one another. In addition, the red and black data points are on the same side for both plots. 
```{r}
plot(wisc.pr$x[, 1], wisc.pr$x[, 3], col = diagnosis, xlab = "PC1", ylab = "PC2")
```

```{r}
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis
library(ggplot2)
ggplot(df) + aes(PC1, PC2, col=diagnosis) + geom_point()
```

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
pve <- pr.var / sum(pr.var)
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```
```{r}
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?
```{r}
component <- wisc.pr$rotation[, 1]
index <- which(names(component) == "concave.points_mean")
concave <- component[index]
concave
```

Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?
```{r}
variance <- which(cumsum(pve) >= 0.80)[1]
variance
```

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method = "complete")
```

Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?
```{r}
plot(wisc.hclust)
abline(h = 4, col = "red", lty = 2)
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
table(wisc.hclust.clusters, diagnosis)
```
Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?
```{r}
compare <- function(k) 
  {
  wisc.hclust.clusters <- cutree(wisc.hclust, k = k)
  table(wisc.hclust.clusters, diagnosis)
  }
cluster_results <- Map(compare, 2:10)
```

Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.
- I like the average linkage because it provides the average distance between the clusters which ultimately creates more balanced clusters.

```{r}
cumulative.variance <- cumsum(pve)
percent90 <- which.max(cumulative.variance >= 0.90)
selected.components <- wisc.pr$x[, 1:percent90]
wisc.pr.hclust <- hclust(dist(selected.components), method = "ward.D2")
```

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```
```{r}
table(grps, diagnosis)
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
```

```{r}
g <- as.factor(grps)
levels(g)
```

```{r}
g <- relevel(g,2)
levels(g)
```

```{r}
plot(wisc.pr$x[,1:2], col=g)
```
Q15. How well does the newly created model with four clusters separate out the two diagnoses?
- The code created 2 cluster assignments to the diagnoses which allows us to see how well the clusters align with the diagnoses.
```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 2)
table(wisc.pr.hclust.clusters, diagnosis)
```
Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.
```{r}
table(wisc.hclust.clusters, diagnosis)
```
```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
wisc.data <- wisc.df[,-1]
wisc.km <- kmeans(scale(wisc.data), centers = 2, nstart = 20)
diagnosis <- as.factor(wisc.df$diagnosis)
table(wisc.km$cluster, diagnosis)
```

Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?
- The analysis procedure that resulted in a clustering model with the best specficity and sensitivity was ward.D2. 

Q18. Which of these new patients should we prioritize for follow up based on your results?
- PC1 should be prioritized because the cluster is more consistent and less random. 
```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
wisc.data <- wisc.df[,-1]
wisc.pr <- prcomp(wisc.data, scale = TRUE)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=1)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

