---
title: "week010"
author: "Leon Tran"
date: "2023-11-10"
output:
  html_document:
    df_print: paged
  word_document: default
---

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy <- read.csv(candy_file, row.names = 1)
head(candy)
```

Q1. How many different candy types are in this dataset?
```{r}
types <- nrow(candy)
types
```

Q2. How many fruity candy types are in the dataset?
```{r}
fruity <- sum(candy$flavor == "fruity")
fruity
```

Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
```{r}
favorite <- "Air Heads"
favorite
winpercent <- candy[favorite, ]$winpercent
winpercent
```

Q4. What is the winpercent value for “Kit Kat”?
```{r}
kitkat <- candy["Kit Kat", ]$winpercent
kitkat
```

Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
```{r}
tootsie <- candy["Tootsie Roll Snack Bars", ]$winpercent
tootsie
```

Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
- No, there are no variables/columns that looks to be on a different scale because all of them fall under range 0 to 1.
Q7. What do you think a zero and one represent for the candy$chocolate column?
- A zero might represent that there is no chocolate in the candy whereas a one might represent the presence of chocolate in the candy.
Q8. Plot a histogram of winpercent values
```{r}
hist(candy$winpercent, main = "Histogram of Winpercent Values", xlab = "Winpercent")
```

Q9. Is the distribution of winpercent values symmetrical?
- The distribution of winpercent values are skewed to the left of the histogram but is relatively symmetrical.
Q10. Is the center of the distribution above or below 50%?
```{r}
mean <- mean(candy$winpercent)
mean
```
- The center of distribution is below 50%.

Q11. On average is chocolate candy higher or lower ranked than fruit candy?
```{r}
chocolate_logical <- as.logical(candy$chocolate)
chocolate_logical
fruit_logical <- as.logical(candy$fruity)
fruit_logical
```
```{r}
winpercent_chocolate <- mean(candy$winpercent[chocolate_logical])
winpercent_chocolate
winpercent_fruity <- mean(candy$winpercent[fruit_logical])
winpercent_fruity
```
- Chocolate is ranked higher than fruit candy

Q12. Is this difference statistically significant?
```{r}
ttest <- t.test(candy$winpercent[chocolate_logical], candy$winpercent[fruit_logical])
ttest
```
- This difference is statistically significant because the p-value is less than .05.

Q13. What are the five least liked candy types in this set?
```{r}
options(repos = c(CRAN = "https://cran.rstudio.com"))
install.packages("dplyr")
library(dplyr)
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy <- read.csv(candy_file, row.names = 1)
least <- candy %>% arrange(winpercent) %>% head(5)
least
```

Q14. What are the top 5 all time favorite candy types out of this set?
```{r}
favorite <- candy %>% arrange(desc(winpercent)) %>% head(5)
favorite
```

Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
install.packages("ggplot2")
library(ggplot2)
```

```{r}
ggplot(candy) + aes(winpercent, rownames(candy)) + geom_bar(stat = "identity", fill = "black") + labs(title = "Candy Ranking Based on Winpercent", x = "Winpercent", y = "Candy Type")
```

Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
ggplot(candy) + 
  aes(x = winpercent, y = reorder(rownames(candy), winpercent)) + geom_bar(stat = "identity", fill = "black") + labs(title = "Candy Ranking Based on Winpercent", x = "Winpercent", y = "Candy Type")
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
ggplot(candy) + aes(winpercent, reorder(rownames(candy),winpercent)) + geom_col(fill=my_cols) 
```

Q17. What is the worst ranked chocolate candy?
- Nik L Nip
Q18. What is the best ranked fruity candy?
- Reeses Peanut Butter cup

```{r}
install.packages("ggrepel")
library(ggrepel)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) + geom_point(col=my_cols) + geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
```{r}
least_expensive <- candy[which.min(candy$pricepercent), ]
least_expensive
```

Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
top5 <- head(candy[ord, c(11, 12)], n = 5)
top5
```
```{r}
leastof5 <- top5[which.min(top5$winpercent), ]
leastof5
```
Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().
```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

```{r}
install.packages("corrplot")
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
- Fruity and chocolate
Q23. Similarly, what two variables are most positively correlated?
- Winpercent and chocolate

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```
```{r}
plot(pca$x[, c(1, 2)], pch = 16, xlab = "PC1", ylab = "PC2")
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
install.packages("plotly")
library(plotly)
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
- Fruity, hard, and pluribus are picked up strongly by PC1 in the positive direction. These results do make sense because they are consistent with the original variables. 